---
title: "About"
author: "Rachel Hubley"
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false

# Improve digit display and NA display for kable tables
options(scipen = 1, knitr.kable.NA = "")

# Load necessary packages
library(tidyverse)
library(ggplot2)
library(plotly)

```


This blog analyzes U.S. presidential inaugural speeches to track changes in language, sentiment, and readability over time. Using the AFINN sentiment lexicon, Flesch Reading Ease formula, and Flesch-Kincaid formula, we examine how emotional tone and speech complexity have shifted from 1789 to 2009.

# Methodology

## Data Collection & Wrangling Description
The inaugural speeches were scraped from [Avalon Project at Yale Law School](https://avalon.law.yale.edu/subject_menus/inaug.asp). 

After scraping, we removed stop words, as well as "fellow", "citizens", and "people" due to repetition.

Additionally, President Van Buren, President Buchanan, President Garfield, and President Coolidge are missing from the dataset due to unavailable transcripts.

### Sentiment Analysis
For the **Sentiment Analysis Over Time**, we found the total sentiment score of each speech using the AFINN lexicon. The AFINN Lexicon assigns identified words with a value from -5 to 5, with negative scores representing negative sentiment and positive scores representing positive sentiment. We then calculated average sentiment score = (total sentiment) / (number of words recognized by AFINN). This is the *average sentiment score* (per word) for each speech, and it accounts for length of speech and allows for fair comparisons across speeches. 

For the **Normalized Bigram Frequency by Era**, we defined six eras: "Revolution & Early Republic," "Expansion, Reform & Civil War," "Reconstruction to WWI," "Interwar & WWII," "Postwar & Civil Rights," and "Post-Watergate to Obama." Then, we got the got the top 8 biwords/bigrams by era with total count greater than 2. Finally, we "normalized" the counts by total bigrams per era, to allow for fair comparisons across eras, since not all eras contain the same number of bigrams.

### Readability Analysis
Two measures of readability were used: Flesch Reading Ease (FRE) and the Flesch-Kincaid Grade Level (FKGL)

The FRE measures the "readability" of a text and assigns a score that correlates to a specific level.
$$
\text{FRE} = 206.835 - 1.015 \times \left( \frac{\text{Total Words}}{\text{Total Sentences}} \right) - 84.6 \times \left( \frac{\text{Total Syllables}}{\text{Total Words}} \right)
$$

- **100 - 90**: 5th Grade
- **90 - 80**: 6th Grade 
- **80 - 70**: 7th Grade 
- **70 - 60**: 8th & 9th Grade 
- **60 - 50**: 10th to 12th Grade 
- **50 - 30**: College
- **30 - 10**: College Graduate
- **10 - 0**: Professional


The FKGL measures the U.S. reading level of a text.
$$
\text{FKGL} = 0.39 \times \left( \frac{\text{Total Words}}{\text{Total Sentences}} \right) + 11.8 \times \left( \frac{\text{Total Syllables}}{\text{Total Words}} \right) - 15.59
$$
For example:

-   A score of **8.0** means the average 8th grader should be able to understand it.
-   A score of **12.0** suggests it matches a high school senior's reading level.

# Limitations
This analysis is limited by the availability and consistency of data. Four presidents (Van Buren, Buchanan, Garfield, Coolidge) were excluded due to missing transcripts. Additionally, sentiment analysis using the AFINN lexicon captures only surface-level emotional tone and may not fully reflect rhetorical nuance. Readability metrics like Flesch-Kincaid and FRE focus on sentence structure and word complexity, but not on conceptual difficulty, historical context, or rhetorical devices. Lastly, clustering results may vary depending on the selected number of clusters and do not always yield interpretable groupings.

# Future Work
Future research could expand the dataset to include more recent inaugural addresses and improve sentiment analysis by incorporating lexicons better suited to political language. Topic modeling could also offer deeper insights into thematic structure. Finally, extending the analysis to other forms of presidential communication, such as State of the Union addresses or campaign speeches, could provide a more comprehensive view of rhetorical trends over time.

# Appendix



```{r}
#| label: fig-sentiment
#| fig-cap: "AFINN Sentiment Scores of Inaugural Speeches"

```


