---
title: "From Washington to Obama: The Evolution of Presidential Speech"
subtitle: "Comparing the Sentiment, Language Use, and Readability of Presidential Inaugural Speeches Over Time"
author: 
- "Rachel Hubley"
- "Sheryl Solorzano"
- "Josh Balderas"
abstract: "From calls for unity to speeches delivered in times of crisis, presidential inaugural addresses show how leaders shape their language to reflect the moment. In this project, we explore how tone, word choice, and readability have shifted from Washington to Obama using sentiment analysis (AFINN) and readability tools like Flesch-Kincaid. We're interested in the patterns — how presidential speech has changed over time, and what that says about leadership, politics, and the public they’re speaking to."
bibliography: bib/library.bib
---

```{r}
#| label: setup
#| include: false

# Improve digit display and NA display for kable tables
options(scipen = 1, knitr.kable.NA = "")

# Load necessary packages
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggplot2)
library(plotly)
library(tidytext)

```

# Sentiment Analysis Over Time

```{r}
#| label: fig-sentiment
#| fig-cap: "AFINN Sentiment Scores of Inaugural Speeches Over Time."

load("data/inaug_words_sentiment.Rdata")

# Plot bar chart
p <- ggplot(inaug_words_sentiment, aes(
  x = date, y = sentiment_avg,
  text = paste("President:", president,
               "<br>Year:", date,
               "<br>Average Sentiment:", round(sentiment_avg, 3)),
  fill = sentiment_avg)) +
  geom_col(show.legend = FALSE) +
  scale_fill_gradient2(low = "#B22234", mid = "#d6d6d6", high = "steelblue", midpoint = 0) + 
  labs(
    x = "Year",
    y = "Average Sentiment"
  ) +
  theme_minimal()

ggplotly(p, tooltip = "text")

```

This interactive bar chart displays the average AFINN sentiment score of each U.S. presidential inaugural address from 1789 to 2009. The chart was created using ggplot2 with a custom blue–gray–red gradient (scale_fill_gradient2()) and made interactive using plotly, allowing readers to hover over each bar for details like president name, year, and sentiment score [@ggplot2gradient; @plotly]. 

Bars are color-coded by sentiment strength, with blue tones representing more positive language and red tones more negative sentiment. This helps visualize how the emotional tone of inaugural speeches has shifted over time.

While sentiment has remained mostly positive throughout history, we observe notably lower average sentiment during the mid-19th century, especially surrounding the Civil War era under Abraham Lincoln. A brief dip also appears in the early 20th century. More recently, sentiment peaked with George H. W. Bush’s 1989 inaugural address — the highest score in the dataset. Since then, sentiment scores have gradually declined, suggesting a rhetorical shift toward more neutral or somber tones in modern speeches.

> Note: Lincoln’s relatively low sentiment score isn’t a reflection of the purpose or urgency of his speech, but likely a result of the AFINN lexicon assigning negative values to many war-related terms. For more detail on how AFINN works and why this matters, see our [About page](about-page.qmd).

While sentiment helps us understand the emotional tone of inaugural speeches, it doesn’t tell us what presidents were actually talking about. To dive deeper into the substance of presidential rhetoric, we analyzed the most common two-word phrases, or bigrams, used across six major historical eras.

# Normalized Bigram Frequency by Era
::: {.panel-tabset}

### Tab 1: 1789–1815 Revolution & Early Republic
```{r}
#| label: fig-freq-era-1
#| fig-cap: "Top Bigrams in 1789-1815 Inaugural Speeches."

load("data/top_biwords_era.Rdata")

filter(top_biwords_era, era == "1789–1815\nRevolution & Early Republic" ) |>
  ggplot(aes(x = reorder(clean_biword, prop), y = prop)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  scale_x_reordered() +
  labs(
    y = "Proportion of All Bigrams",
    x = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 14),
    axis.text.x = element_text(size = 10)
  )

```


### Tab 2: 1816–1865 Expansion, Reform & Civil War
```{r}
#| label: fig-freq-era-2
#| fig-cap: "Top Bigrams in 1816–1865 Inaugural Speeches."


filter(top_biwords_era, era == "1816–1865\nExpansion, Reform & Civil War") |>
  ggplot(aes(x = reorder(clean_biword, prop), y = prop)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  scale_x_reordered() +
  labs(
    y = "Proportion of All Bigrams",
    x = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 14),
    axis.text.x = element_text(size = 10)
  )

```


### Tab 3: 1866–1919 Reconstruction to WWI
```{r}
#| label: fig-freq-era-3
#| fig-cap: "Top Bigrams in 1866-1919 Inaugural Speeches."

filter(top_biwords_era, era == "1866–1919\nReconstruction to WWI") |>
  ggplot(aes(x = reorder(clean_biword, prop), y = prop)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  scale_x_reordered() +
  labs(
    y = "Proportion of All Bigrams",
    x = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 14),
    axis.text.x = element_text(size = 10)
  )


```


### Tab 4: 1920–1945 Interwar & WWII
```{r}
#| label: fig-freq-era-4
#| fig-cap: "Top Bigrams in 1920-1945 Inaugural Speeches."

filter(top_biwords_era, era == "1920–1945\nInterwar & WWII") |>
  ggplot(aes(x = reorder(clean_biword, prop), y = prop)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  scale_x_reordered() +
  labs(
    y = "Proportion of All Bigrams",
    x = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 14),
    axis.text.x = element_text(size = 10)
  )

```


### Tab 5: 1946–1969 Postwar & Civil Rights
```{r}
#| label: fig-freq-era-5
#| fig-cap: "Top Bigrams in 1946-1969 Inaugural Speeches."

filter(top_biwords_era, era == "1946–1969\nPostwar & Civil Rights") |>
  ggplot(aes(x = reorder(clean_biword, prop), y = prop)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  scale_x_reordered() +
  labs(
    y = "Proportion of All Bigrams",
    x = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 14),
    axis.text.x = element_text(size = 10)
  )

```


### Tab 6: 1970–2009 Post-Watergate to Obama
```{r}
#| label: fig-freq-era-6
#| fig-cap: "Top Bigrams in 1970-2009 Inaugural Speeches."


filter(top_biwords_era, era == "1970–2009\nPost-Watergate to Obama") |>
  ggplot(aes(x = reorder(clean_biword, prop), y = prop)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  scale_x_reordered() +
  labs(
    y = "Proportion of All Bigrams",
    x = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 14),
    axis.text.x = element_text(size = 10)
  )


```

:::

*Bigrams marked with an asterisk (*) appear in more than one era.

This faceted bar chart highlights the most frequently used bigrams (two-word phrases) in inaugural speeches across six major historical eras. Frequencies are normalized to account for differences in era length and speech length, so bars represent the proportion of all bigrams within that era. Red bars indicate bigrams that appear in multiple eras, suggesting persistent rhetorical themes (such as "federal government" or "foreign nations").

A key trend revealed here is how certain bigrams like “federal government” and “foreign nations” recur across multiple eras, showing the enduring importance of governance and international relations. However, top bigrams also shift to reflect evolving national concerns: for example, “united nations” emerges as the most dominant phrase during the Postwar & Civil Rights era (1946–1969), while “god bless” and “vice president” gain prominence in the Post-Watergate to Obama era (1970–2009). This suggests a rhetorical turn toward both global institutions and more personal or moral language in the modern period.

A key trend revealed here is how certain bigrams like “federal government” and “foreign nations” recur across multiple eras, showing the enduring importance of governance and international relations. However, the top bigrams also shift to reflect evolving national concerns:

- In the Reconstruction to WWI era (1866–1919), bigrams like “civil service” and “public expenditures” reflect growing attention to internal administration and government reform.

- During the Interwar & WWII period (1920–1945), phrases such as “eighteenth amendment” and “illegal liquor" reference Prohibition.

- In the Postwar & Civil Rights era (1946–1969), the emergence of phrases like “armed attack” and “free nations” indicates heightened attention to Cold War tensions and the defense of global democracy.

- Finally, in the Post-Watergate to Obama era (1970–2009), we see phrases like “nuclear weapons” and “god bless,” suggesting a shift toward national security concerns, as well with moral and religious appeals in presidential rhetoric. This era shows a mix of national security concerns and attempts to bring people together using faith during times of global and domestic uncertainty.

These changes underscore how presidential language not only reflects shared values, but also adapts to shifting historical and political priorities.

Because each panel shows a different historical era, bars needed to be reordered within each facet. This was accomplished using reorder_within() from the tidytext package, following the solution described by Julia Silge [@reorderwithin]. Facet label formatting was adapted using advice from Kassambara [@facetlabels], and the scales = "free_y" setting was enabled to allow y-axes to scale independently [@free_scales].


# Readability Analysis

## Introduction

Readability analysis helps us understand how easy (or difficult) a text is to read. Two common measures are the Flesch Reading Ease (FRE) and Flesch-Kincaid Grade Level (FKGL)

## Understanding the relationship: Flesch Reading Ease (FRE) & Flesch-Kincaid Grading Level (FKGL)

To get a better understanding of the relationship between the Flesch Reading Ease (FRE) and the Flesch-Kincaid Grade Level (FKGL), it is helpful to plot them together on a scatterplot. 

```{r}
#| label: fig-FKGL-vs-FRE
#| fig-cap: "This chart displays the inverse relationship between the Flesch-Kincaid and Flesch Reading Ease formulas."

load("data/inaugural_speeches_master.Rdata")
a <- ggplot(inaug_speeches_master, aes(x = Flesch, y = Flesch.Kincaid)) +
  geom_point() +
  geom_smooth( method = "loess", se = TRUE) +
  scale_x_continuous(breaks = scales:: breaks_width(10), minor_breaks = NULL) +
  labs(title = "FRE vs FKGL",
    y = "Flesch-Kincaid Grade Level",
       x = "Flesch Reading Ease") +
  theme_minimal()


ggplotly(a, tooltip = "text")

```


The FRE and the FKGL have an inverse relationship because they are based on similar textual features—sentence length and word complexity—but use opposing scales to express readability.


$$
\text{FRE} = 206.835 - 1.015 \times \left( \frac{\text{Total Words}}{\text{Total Sentences}} \right) - 84.6 \times \left( \frac{\text{Total Syllables}}{\text{Total Words}} \right)
$$
$$
\text{FKGL} = 0.39 \times \left( \frac{\text{Total Words}}{\text{Total Sentences}} \right) + 11.8 \times \left( \frac{\text{Total Syllables}}{\text{Total Words}} \right) - 15.59
$$

As sentence length and syllables per word increase, FRE scores decrease (the text becomes harder to read), while FKGL scores increase (the grade level rises). This opposing behavior is why they have an inverse relationship: when one goes up, the other tends to go down.


## Flesch Reading Ease (FRE)

$$
\text{FRE} = 206.835 - 1.015 \times \left( \frac{\text{Total Words}}{\text{Total Sentences}} \right) - 84.6 \times \left( \frac{\text{Total Syllables}}{\text{Total Words}} \right)
$$

This score rates text on a 0 to 100 scale. A higher score means the text is easier to read.

- **100 - 90**: 5th Grade
- **90 - 80**: 6th Grade 
- **80 - 70**: 7th Grade 
- **70 - 60**: 8th & 9th Grade 
- **60 - 50**: 10th to 12th Grade 
- **50 - 30**: College
- **30 - 10**: College Graduate
- **10 - 0**: Professional


### FRE vs. Sentiment

To better understand the relationship between writing style and emotional tone in the presidential inaugural addresses, we compare the FRE scores with sentiment scores calculated for each speech.

The FRE measures how easy a text is to read, while the sentiment score captures the overall positivity or negativity of the language. By analyzing these two metrics together, we can apply clustering techniques to group speeches with similar readability and sentiment profiles.

```{r}
#| label: fig-Flesch_Sentiment
#| fig-cap: "Flesch Reading Ease vs. Sentiment Scores from inaugural speech."
p <- ggplot(inaug_speeches_master, aes(y = Flesch, 
                                       x = sentiment, 
                                       text = paste("President:", 
                                                    president,
                                                    "<br>Year:", 
                                                    date,
                                                    "<br>Sentiment Score:", 
                                                    sentiment,
                                                    "<br>Flesch Score:",
                                                    Flesch))
            ) +
  scale_color_manual(values = c("red", "blue", "green", "purple", "orange", "pink")) +
  geom_point(aes(color = factor(flesch_sentiment_cluster))) +
  scale_x_continuous(breaks = scales:: breaks_width(25), minor_breaks = NULL) +
  labs(y = "Flesch Reading Level",
       x = "Sentiment Score",
       color = "Cluster") +
  theme_minimal()

ggplotly(p, tooltip = "text")

```


Using the clustering analysis, some light interpretation can be made from the groupings.

-   **Cluster 1**: Presidents have average sentiment score with low readability.

-   **Cluster 2**: Presidents have higher sentiment score with low readability.

-   **Cluster 3**: Presidents have higher sentiment score with high readability.

-   **Cluster 4**: Presidents have low sentiment scores with high readability.

-   **Cluster 5**: Presidents on the lower end of the sentiment score spectrum. Readability ranges from standard to easy.

-   **Cluster 6**: Presidents in this cluster have the highest sentiment score spectrum. Readability ranges from 30-40.


## Flesch-Kincaid Grade Level (FKGL)

$$
\text{FKGL} = 0.39 \times \left( \frac{\text{Total Words}}{\text{Total Sentences}} \right) + 11.8 \times \left( \frac{\text{Total Syllables}}{\text{Total Words}} \right) - 15.59
$$

This score estimates the U.S. school grade level needed to understand the text. Due to this formulas construction, the score does not have an upper bound. 

For example:

-   A score of **8.0** means the average 8th grader should be able to understand it.
-   A score of **12.0** suggests it matches a high school senior's reading level.

### FKGL vs. Sentiment

In addition to examining readability through the Flesch score, we also analyze the FKGL in relation to sentiment scores for each inaugural address. By comparing FKGL to sentiment scores, we explore how the sophistication of presidential language interacts with the emotional tone conveyed while using clustering methods.

```{r}
#| label: fig-Flesch-Kincaid-vs-Sentiment
#| fig-cap: "This chart displays Flesch-Kincaid and sentiment scores of president speech values."

p <- ggplot(inaug_speeches_master, aes(y = Flesch.Kincaid, x = sentiment,
                                        text = paste("President:", 
                                                    president,
                                                    "<br>Year:", 
                                                    date,
                                                    "<br>Sentiment Score:", 
                                                    sentiment,
                                                    "<br>FK Score:",
                                                    Flesch.Kincaid))) +
  scale_color_manual(values = c("red", "blue", "green", "purple", "orange", "pink")) +
  geom_point(aes(color = factor(flesch_kin_sentiment_cluster))) +
  scale_x_continuous(breaks = scales:: breaks_width(25), minor_breaks = NULL) +
  labs(y = "Flesch-Kincaid Grade Level",
       x = "Sentiment Score",
       color = "Cluster") +
  theme_minimal()

ggplotly(p, tooltip = "text")
```

In this example, clustering analysis using FKGL vs Sentiment values does not yield any meaningful results. Clustering appears to be driven entirely by the sentiment axis and form vertical "layers" that reflect sentiment values alone. 

This suggests that the algorithm is not detecting any significant structure in the Flesch-Kincaid axis, and as a result, the clustering lacks any interpretive value.

### FKGL vs. Time

We also examine the relationship between the FKGL and time to track how the complexity of presidential inaugural speeches has changed. Overall, we find a clear trend that speech complexity has decreased over time, with more recent presidents using simpler, more accessible language compared to earlier eras.

```{r}
#| label: fig-Flesh-Kincaid-over-time
#| fig-cap: "This chart displays Flesch-Kincaid over time illustrating the complexity of inaugural addresses."

p <- ggplot(inaug_speeches_master, aes(x = date, y = Flesch.Kincaid))+
  geom_point() +
  geom_smooth( method = "loess") +
  scale_x_continuous(breaks = scales:: breaks_width(25), minor_breaks = NULL) +
  labs(y = "Flesch-Kincaid Grade Level",
       x = "Year") +
  theme_minimal()

ggplotly(p, tooltip = "text")

```

Possible explanations include:

-   Initially, the audience of the inaugural speech was composed of congressional leaders.
-   The growing emphasis on mass communication, as presidents began speaking to a broader and more diverse audience.
-   The influence of modern media, which favors shorter, clearer messaging that resonates quickly with the public.


```{r}
#| label: fig-Top-5-presidents-with-lowest-scores
#| fig-cap: "Top 5 presidents with lowest Flesch-Kincaid scores"

# Loading the data
load("data/top5_lowpress.Rdata")


# table displaying the top 5 presidents with the highest Flesch-Kincaid score 
 top5_lowpress |>
  kable(booktabs = TRUE, 
        digits = 2,
        col.names = c("President", "Year", "Flesch-Kincaid Score")) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center"
  ) |>
  row_spec(0, bold = TRUE, color = "white", background = "#000080") |>
  column_spec(2, color = "black", background = "#ffffff") |>
  column_spec(3, color = "black", background = "#ffffff") |>
  add_header_above(c("Top Five Presidents with Lowest \n Flesch-Kincaid Scores" = 3),
                   background = "#CD2626", color = "white", bold = TRUE)
 
```


```{r}
#| label: fig-Top-5-presidents-with-highest-scores
#| fig-cap: "Top 5 presidents with the highest Flesch-Kincaid scores"

# Loading the data
load("data/top5_highpress.Rdata")

# table displaying the top 5 presidents with the lowest Flesch-Kincaid scores 
  top5_highpress |>
  kable(booktabs = TRUE, 
        digits = 2,
        col.names = c("President", "Year", "Flesch-Kincaid Score")) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center"
  ) |>
  row_spec(0, bold = TRUE, color = "white", background = "#000080") |>
  column_spec(2, color = "black", background = "#ffffff") |>
  column_spec(3, color = "black", background = "#ffffff") |>
  add_header_above(c("Top Five Presidents with Highest \n Flesch-Kincaid Scores" = 3),
                   background = "#CD2626", color = "white", bold = TRUE)
```

Figure 10a displays a table ranking the five presidents with the lowest Flesch-Kincaid Grade Level scores between 1789 and 2009. Similarly, Figure 10b displays a table ranking the five presidents with the highest Flesch-Kincaid Grade Level scores. By looking at the years, we can see how the highest scores correspond to the presidents in the first two centuries while the lowest scores correspond to the presidents in the more recent centuries. This indicates that the language used in inaugural speeches has become simpler and more accessible over time. The tables' feature and other attributes were accomplished from the `kabbleExtra` package by following the documentation provided by CRAN [@kableTable].


# Sentiment and Flesch-Kincaid Scores Change among Presidents that Served Two Terms


```{r}
#| label: fig-sentiment-two-terms
#| fig-cap: "Sentiment score change from a president's first term to their second term"

#loading the data
load("data/sec_terms_sent_fk.Rdata")


# Inaugural Address Complexity Change from First to Second Term for presidents that served two terms
q <- ggplot(sec_terms_sent_fk, aes(x = sent_first, y = sent_second,
                                   text = paste("President:", 
                                                    president,
                                                    "<br> 1st Term Sentiment Score:",
                                                    round(sent_first, 3),
                                                "<br> 2nd Term Sentiment Score:",
                                                round(sent_second, 3) ) )) +
  geom_point(aes(color = president, size = 0.1)) + 
  labs(title = "Inaugural Address Sentiment Score: First Term vs. Second Term",
       x = "First Term Sentiment Avg",
       y = "Second Term Sentiment Avg") +
  theme_minimal()

# Removing the legend 
ggplotly(q, tooltip = "text") |>
  layout(showlegend = FALSE)

```


```{r}
#| label: fig-FK-two-terms-
#| fig-cap: "Flesch-Kincaid score change from a president's first term to their second term"
# Inaugural Address Complexity Change from First to Second Term for presidents that served two terms
q <- ggplot(sec_terms_sent_fk, aes(x = fk_first, y = fk_second,
                                   text = paste("President:", 
                                                    president,
                                                    "<br> 1st Term FK Score:",
                                                    fk_first,
                                                "<br> 2nd Term FK Score:",
                                                fk_second))) +
  geom_point(aes(color = president, size = 0.1)) + 
  labs(title = "Inaugural Address Complexity: First Term vs. Second Term",
       x = "First Term Flesch-Kincaid Score",
       y = "Second Term Flesch-Kincaid Score") +
  theme_minimal()

# Removing the legend
ggplotly(q, tooltip = "text") |>
  layout(showlegend = FALSE)

```

## Difference Score

```{r}
#| label: fig-sentiment-FK-across-terms
#| fig-cap: "Sentiment score and Flesch-Kincaid Scores change across different terms of presidents that served two terms"

# loading the data
load("data/sec_terms_diff.Rdata")


# calculating the differences from the first term to the second term of presidents in terms of both sentiment score and flesch-kincaid scores

sec_terms_diff <- sec_terms_sent_fk |> 
  mutate(
    sentiment_diff = sent_second - sent_first,
    fk_diff = fk_second - fk_first
  ) |>
  select(president, sentiment_diff, fk_diff)

 
 # table displaying the calculated difference in sentiment scores and Flesch-Kincaid Scores
  sec_terms_diff |>
  kable(booktabs = TRUE, 
        digits = 2,
        col.names = c("President", "Sentiment Score Difference", "Flesch-Kincaid Score Difference")) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center"
  ) |>
  row_spec(0, bold = TRUE, color = "white", background = "#000080") |>
  column_spec(2, color = "black", background = "#ffffff") |>
  column_spec(3, color = "black", background = "#ffffff") |>
  add_header_above(c("Differences Between Terms" = 3),
                   background = "#CD2626", color = "white", bold = TRUE)
```

Figure 13 displays an interactive scatter plot showing the sentiment scores of U.S. presidents’ first and second inaugural addresses for those who served two terms. Similarly, Figure 6 shows presidents' first term Flesch-Kincaid scores and second term Flesch-Kincaid scores. By hovering over each data point, the user can view the president's name as well as their respective scores. Each president is colored differently.

By examining the sentiment and Flesch-Kincaid scores across different terms, we can explore how presidential leadership and political rhetoric evolve after serving a first term. This is important because it may reveal a correlation between a president’s experience and changes in language complexity or emotional tone

To see this more clearly, Figure 13a displays a table showing the changes in sentiment scores and Flesch-Kincaid scores for all presidents that served two terms. This will facilitate determining whether their tone became more positive or negative and whether their language became more complex or simpler across different terms.

Sentiment Score difference:

 - **Positive Difference Value**: The second-term inaugural speech was more postive than their first term inaugural speech.
 - **Negative Difference Value**: The second-term inaugural speech was less postive or more negative than the first term inaugural speech.
 - **Zero or close to Zero Difference Value**: The sentiment/tone of the inaugural speech remained pretty consistent.
 
 
 Flesch-Kincaid Score Difference:
 
 - **Positive Difference Value**: The second-term inaugural speech was more complex than the first term inaugural speech.
 - **Negative Difference Value**: The second-term inaugural speech was simpler and easier to understand than the first term inaugural speech.
 - **Zero or close to Zero Difference Value**: The complexity of the inaugural speech remained pretty consistent.
 
