---
title: "Sentiment Analysis of Inaugural Speeches"
subtitle: "Comparing the Sentiment of Presidential Speeches Over Time"
author: 
- "Rachel Hubley"
- "Sheryl Solorzano"
- "Josh Balderas"
abstract: "This blog focuses on analyzing the inaugural speeches of U.S. presidents to examine how the language, sentiment used, and readability of presidential addresses have changed over time. By calculating readability scores using the Flesch-Kincaid formula and analyzing sentiment with the AFINN lexicon, we explore whether speeches from 1789 to 2009 have become easier or harder to understand, and how their emotional tone has evolved. Please note that President Van Buren, President Buchanan, President Garfield, and President Coolidge are missing from the dataset due to unavailable transcripts."
bibliography: bib/library.bib
---

```{r}
#| label: setup
#| include: false

# Improve digit display and NA display for kable tables
options(scipen = 1, knitr.kable.NA = "")

# Load necessary packages
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggplot2)
library(plotly)
library(tidytext)

```

# Sentiment Analysis Over Time

```{r}
#| label: fig-sentiment
#| fig-cap: "AFINN Sentiment Scores of Inaugural Speeches Over Time."

load("data/inaug_words_sentiment.Rdata")

# Plot bar chart
p <- ggplot(inaug_words_sentiment, aes(
  x = date, y = sentiment_avg,
  text = paste("President:", president,
               "<br>Year:", date,
               "<br>Average Sentiment:", round(sentiment_avg, 3)),
  fill = sentiment_avg)) +
  geom_col(show.legend = FALSE) +
  scale_fill_gradient2(low = "#B22234", mid = "#d6d6d6", high = "steelblue", midpoint = 0) + 
  labs(
    title = "AFINN Sentiment Scores of Inaugural Speeches",
    x = "Year",
    y = "Average Sentiment"
  ) +
  theme_minimal()

ggplotly(p, tooltip = "text")

```

This interactive bar chart displays the average AFINN sentiment score of each U.S. presidential inaugural address over time. Bars are color-coded by sentiment strength, with greener bars indicating more positive language and blue tones indicating more negative sentiment. Hovering over each bar reveals additional details, including the president’s name, year of the address, and exact sentiment score. 

While sentiment has remained mostly positive throughout history, we observe notably lower average sentiment during the mid-19th century, especially surrounding the Civil War era under Abraham Lincoln. A brief dip also appears in the early 20th century. More recently, sentiment peaked with George H. W. Bush’s 1989 inaugural address, which had the highest score in the dataset. Since then, average sentiment scores have gradually declined across subsequent presidencies, suggesting a rhetorical shift toward more neutral or somber tones in modern speeches.

The first figure uses an interactive plotly chart to allow readers to explore sentiment scores by hovering over each bar [@plotly]. The color gradient, from blue to gray to red, was created using scale_fill_gradient2() in ggplot2 [@ggplot2gradient].

# Normalized Bigram Frequency by Era

```{r}
#| label: fig-freq-era
#| fig-cap: "Top Biwords in Inaugural Speeches by Era."
#| layout-ncol: 1
#| fig-width: 17
#| fig-height: 10
#| out-width: 97%
#| fig-align: center


load("data/top_biwords_era.Rdata")
  
# ggplot(top_biwords_era, aes(x = reorder_within(biword, total, era), y = total)) +
#   geom_col(fill = "steelblue", show.legend = FALSE) +
#   coord_flip() +
#   facet_wrap(~ era, scales = "free_y") +
#   scale_x_reordered() +
#   scale_y_continuous(breaks = 0:12) 
#   labs(y = "Bigram Count") +
#   theme_void() +
#   theme(
#     strip.text = element_text(face = "bold", size = 18),
#     axis.text = element_text(size = 14),
#     plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
#     plot.caption = element_text(size = 10, hjust = 0.5)
#   )


# ggplot(top_biwords_era, aes(x = biword, y = total, fill = color)) +
#   geom_col(show.legend = FALSE) +
#   coord_flip() +
#   facet_wrap(~ era, scales = "free_y") +
#   scale_x_reordered() +  
#   scale_y_continuous(breaks = 0:12) +
#   scale_fill_identity() +            
#   labs(
#     x = NULL,
#     y = "Bigram Count"
#   ) +
#   theme_minimal() +
#   theme(
#     strip.text = element_text(face = "bold", size = 18),
#     axis.text = element_text(size = 14),
#     plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
#     plot.caption = element_text(size = 10, hjust = 0.5)
#   )


ggplot(top_biwords_era, aes(x = biword, y = prop, fill = color)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ era, scales = "free_y") +
  scale_x_reordered() +
  scale_fill_identity() +
  labs(
    y = "Proportion of All Bigrams",
    x = NULL
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(margin = margin(t = 8, b = 8), size = 16, face = "bold"),
    axis.title.x = element_text(size = 20),
    axis.text = element_text(size = 14),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.caption = element_text(size = 10, hjust = 0.5),
    plot.margin = margin(t = 40, r = 20, b = 10, l = 10)
  )


```

This faceted bar chart highlights the most frequently used bigrams (two-word phrases) in inaugural speeches across six major historical eras. Frequencies are normalized to account for differences in era length and speech length, so bars represent the proportion of all bigrams within that era. Red bars indicate bigrams that appear in multiple eras, suggesting persistent rhetorical themes (such as "federal government" or "foreign nations").

A key trend revealed here is how certain bigrams like “federal government” and “foreign nations” recur across multiple eras, showing the enduring importance of governance and international relations. However, top bigrams also shift to reflect evolving national concerns: for example, “united nations” emerges as the most dominant phrase during the Postwar & Civil Rights era (1946–1969), while “god bless” and “vice president” gain prominence in the Post-Watergate to Obama era (1970–2009). This suggests a rhetorical turn toward both global institutions and more personal or moral language in the modern period.


A key trend revealed here is how certain bigrams like “federal government” and “foreign nations” recur across multiple eras, showing the enduring importance of governance and international relations. However, the top bigrams also shift to reflect evolving national concerns:

- In the Reconstruction to WWI era (1866–1919), bigrams like “civil service” and “public expenditures” reflect growing attention to internal administration and government reform.

- During the Interwar & WWII period (1920–1945), phrases such as “eighteenth amendment” and “illegal liquor" reference Prohibition.

- In the Postwar & Civil Rights era (1946–1969), the emergence of phrases like “armed attack” and “free nations” indicates heightened attention to Cold War tensions and the defense of global democracy.

- Finally, in the Post-Watergate to Obama era (1970–2009), we see phrases like “nuclear weapons” and “god bless,” suggesting a shift toward national security concerns, as well with moral and religious appeals in presidential rhetoric. This era shows a mix of national security concerns and attempts to bring people together using faith during times of global and domestic uncertainty.

These changes underscore how presidential language not only reflects shared values, but also adapts to shifting historical and political priorities.

Because each panel shows a different historical era, bars needed to be reordered within each facet. This was accomplished using reorder_within() from the tidytext package, following the solution described by Julia Silge [@reorderwithin]. Facet label formatting was adapted using advice from Kassambara [@facetlabels], and the scales = "free_y" setting was enabled to allow y-axes to scale independently [@free_scales].


# Mean Sentence Length over time 

```{r}
#| label: Mean Sentence Length over time
#| fig-cap: "Figure 4: This bar char displays the sentence mean in each president's inaugural speech over time"

# loading the data
load("data/inaug_speeches_master")

load("data/inaugural_speeches_master.Rdata")

ms <- ggplot(inaug_speeches_master, aes(
  x = date, y = meanSentenceLength,
  text = paste("President:", president,
               "<br>Year:", date,
               "<br>Mean Sentence Length:", round(meanSentenceLength, 3)),
  fill = meanSentenceLength)) +
  geom_col(show.legend = FALSE) +
  scale_fill_gradient2(low = "#d61106", mid = "#d6d6d6", high = "#05853c", midpoint = 0) + 
  labs(
    title = "Mean Sentence Length of Inaugural Speeches",
    x = "Year",
    y = "Mean Sentence Length"
  ) +
  theme_minimal()

ggplotly(ms, tooltip = "text")

```


[summary of the figure]


# Flesch-Kincaid Scores over Time

```{r}
#| label: Flesch-Kincaid Scores over time
#| fig-cap: "Figure 5: This bar char displays the complexity of presidents' inaugural speeches with their respective Flesch-Kincaid scores over time"


fk <- ggplot(inaug_speeches_master, aes(
  x = date, y = Flesch.Kincaid,
  text = paste("President:", president,
               "<br>Year:", date,
               "<br>Flesch.Kincaid:", round(Flesch.Kincaid, 3)),
  fill = Flesch.Kincaid)) +
  geom_col(show.legend = FALSE) +
  scale_fill_gradient2(low = "#d61106", mid = "#d6d6d6", high = "#05853c", midpoint = 0) + 
  labs(
    title = "Flesch-Kincaid Score of Inaugural Speeches",
    x = "Year",
    y = "Flesch-Kincaid Score"
  ) +
  theme_minimal()

ggplotly(fk, tooltip = "text")
```

[summary of the figure]


```{r}
#| label: Top 5 presidents with lowest scores
#| Fig-cap: "...."

# Loading the data
load("data/top5_lowpress.Rdata")


# table displaying the top 5 presidents with the highest Flesch-Kincaid score 
 top5_lowpress |>
  kable(booktabs = TRUE, 
        digits = 2,
        col.names = c("President", "Year", "Flesch-Kincaid Score")) |>
  kable_styling()
 
  
```

{summary} George Bush is the president with the lowest Flesch-Kincaid score.


```{r}
#| label: Top 5 presidents with highest scores
#| Fig-cap: "..."
# Loading the data
load("data/top5_highpress.Rdata")

# table displaying the top 5 presidents with the lowest Flesch-Kincaid scores 
  top5_highpress |>
  kable(booktabs = TRUE, 
        digits = 2,
        col.names = c("President", "Year", "Flesch-Kincaid Score")) |>
  kable_styling()
```


[summary] John Adams is the president with the highest Flesch-Kincaic score.






# Sentiment Scores and Flesch-Kincaid Scores Change among Presidents that Served Two Terms

## First Term Sentiment Score vs. Second Term Sentiment Score

```{r}
#| Fig-cap: "..."

#loading the data
load("data/sec_terms_sent_fk.Rdata")


# Inaugural Address Complexity Change from First to Second Term for presidents that served two terms
q <- ggplot(sec_terms_sent_fk, aes(x = sent_first, y = sent_second,
                                   text = paste("President:", 
                                                    president,
                                                    "<br> 1st Term Sentiment Score:",
                                                    round(sent_first, 3),
                                                "<br> 2nd Term Sentiment Score:",
                                                round(sent_second, 3) ) )) +
  geom_point(aes(color = president, size = 0.1)) + 
  labs(title = "Inaugural Address Sentiment Change from First to Second Term",
       x = "First Term Sentiment Avg",
       y = "Second Term Sentiment Avg") +
  theme_minimal()

# Removing the legend 
ggplotly(q, tooltip = "text") |>
  layout(showlegend = FALSE)

```

## First Term Flesch-Kincaid Score vs. Second Term Flesch-Kincaid Score

```{r}
# Inaugural Address Complexity Change from First to Second Term for presidents that served two terms
q <- ggplot(sec_terms_sent_fk, aes(x = fk_first, y = fk_second,
                                   text = paste("President:", 
                                                    president,
                                                    "<br> 1st Term FK Score:",
                                                    fk_first,
                                                "<br> 2nd Term FK Score:",
                                                fk_second))) +
  geom_point(aes(color = president, size = 0.1)) + 
  labs(title = "Inaugural Address Complexity Change from First to Second Term",
       x = "First Term Flesch-Kincaid Score",
       y = "Second Term Flesch-Kincaid Score") +
  theme_minimal()

# Removing the legend
ggplotly(q, tooltip = "text") |>
  layout(showlegend = FALSE)

```

## Difference Score

```{r}
# loading the data
load("data/sec_terms_diff.Rdata")

# calculating the differences from the first term to the second term of presidents in terms of both sentiment score and flesch-kincaid scores

sec_terms_diff <- sec_terms_sent_fk |> 
  mutate(
    sentiment_diff = sent_second - sent_first,
    fk_diff = fk_second - fk_first
  ) |>
  select(president, sentiment_diff, fk_diff)

 
 # table displaying the calculated difference in sentiment scores and Flesch-Kincaid Scores
  sec_terms_diff |>
  kable(booktabs = TRUE, 
        digits = 2,
        col.names = c("President", "Sentiment Score Difference", "Flesch-Kincaid Score Difference")) |>
  kable_styling()
```

Sentiment Score difference:

 - **Positive Difference Value**: The second-term inaugural speech was more postive than the first term that preseident served
 - **Negative Difference Value**: The second-term inaugural speech was less postive or more negative than the first term speech of that president
 - **Zero or close to Zero Difference Value**: The sentiment/tone remained pretty consistent 
 
 
 Flesch-Kincaid Score Difference:
 
 - **Positive Difference Value**: The second-term inaugural speech was more complex than the first term that preseident served
 - **Negative Difference Value**: The second-term inaugural speech was simpler and easier to understand than the first term speech of that president
 - **Zero or close to Zero Difference Value**: The complexity of the inaugural speech remained pretty consistent
 
 
## Mean Sentence Length vs. Sentiment
 
```{r, warning = FALSE}
#| label: Plotting Mean Sentence Length vs. Sentiment

load("data/msl_cluster_centers")
load("data/msl_clustered")


# Interactive plot with plotly
MS_s <- ggplot(msl_clustered, aes(x = meanSentenceLength, y = sentiment_avg)) +
  geom_point(aes(color = cluster_2, 
                 text = paste("President:", president,
                              "<br>Sentiment:", sentiment_avg,
                              "<br>Mean Sentence Length:", meanSentenceLength,
                              "<br>Cluster:", cluster_2)), 
             alpha = 0.6) +
  geom_label(data = msl_cluster_centers, 
             aes(label = cluster_2, 
                 text = paste("Cluster Center:", cluster_2)), 
             size = 3) +
  labs(y = "Average Sentiment Score",
       x = "Mean Sentence Length",
       title = "Inaugural Address Analysis",
       caption = "Average Sentiment Score vs Mean Sentence Length",
       color = "Cluster") +
  theme_minimal()

ggplotly(MS_s, tooltip = "text")


```
 
 
 
 
 
 
 
 

# Readability Analysis

## Introduction

Readability analysis helps us understand how easy (or difficult) a text is to read. Two common measures are the Flesch Reading Ease (FRE) and Flesch-Kincaid Grade Level (FKGL)

## Flesch Reading Ease (FRE)

$$
\text{FRE} = 206.835 - 1.015 \times \left( \frac{\text{Total Words}}{\text{Total Sentences}} \right) - 84.6 \times \left( \frac{\text{Total Syllables}}{\text{Total Words}} \right)
$$

This score rates text on a 0 to 100 scale. A higher score means the text is easier to read.

- **100 - 90**: 5th Grade
- **90 - 80**: 6th Grade 
- **80 - 70**: 7th Grade 
- **70 - 60**: 8th & 9th Grade 
- **60 - 50**: 10th to 12th Grade 
- **50 - 30**: College
- **30 - 10**: College Graduate
- **10 - 0**: Professional


### FRE vs. Sentiment

To better understand the relationship between writing style and emotional tone in the presidential inaugural addresses, we compare the Flesch Reading Ease scores with sentiment scores calculated for each speech.

The Flesch score measures how easy a text is to read, while the sentiment score captures the overall positivity or negativity of the language. By analyzing these two metrics together, we can apply clustering techniques to group speeches with similar readability and sentiment profiles. These clusters help us identify distinct subgroups of inaugural addresses --- for example, groups of speeches that are both highly readable and optimistic, or more complex and somber --- offering new insights into how presidents have tailored their language to the moment in history.

```{r}
#| label: Plotting-Flesch_Sentiment
#| fig-cap: "Figure [?]: This chart displays Flesch reading ease and sentiment scores of president speech values."
p <- ggplot(inaug_speeches_master, aes(y = Flesch, 
                                       x = sentiment, 
                                       text = paste("President:", 
                                                    president,
                                                    "<br>Year:", 
                                                    date,
                                                    "<br>Sentiment Score:", 
                                                    sentiment,
                                                    "<br>Flesch Score:",
                                                    Flesch))
            ) +
  scale_color_manual(values = c("red", "blue", "green", "purple", "orange", "pink")) +
  geom_point(aes(color = factor(flesch_sentiment_cluster))) +
  scale_x_continuous(breaks = scales:: breaks_width(25), minor_breaks = NULL) +
  labs(y = "Flesch Reading Level",
       x = "Sentiment Score",
       title = "Inaugural Address Analysis",
       subtitle = "Flesch Reading Level vs Sentiment Score",
       color = "Cluster") +
  theme_minimal()

ggplotly(p, tooltip = "text")

```


Using the clustering analysis, some light interpretation can be made from the groupings.

-   **Cluster 1**: Presidents have average sentiment score with low readability.

-   **Cluster 2**: Presidents have higher sentiment score with low readability.

-   **Cluster 3**: Presidents have higher sentiment score with high readability.

-   **Cluster 4**: Presidents have low sentiment scores with high readability.

-   **Cluster 5**: Presidents on the lower end of the sentiment score spectrum. Readability ranges from standard to easy.

-   **Cluster 6**: Presidents in this cluster have the highest sentiment score spectrum. Readability ranges from 30-40.


## Flesch-Kincaid Grade Level (FKGL)

$$
\text{FKGL} = 0.39 \times \left( \frac{\text{Total Words}}{\text{Total Sentences}} \right) + 11.8 \times \left( \frac{\text{Total Syllables}}{\text{Total Words}} \right) - 15.59
$$

This score estimates the U.S. school grade level needed to understand the text and corresponds with a U.S. grade level. Due to this formulas construction, the score does not have an upper bound. 

For example:

-   A score of **8.0** means the average 8th grader should be able to understand it.
-   A score of **12.0** suggests it matches a high school senior's reading level.

### FKGL vs. Sentiment

In addition to examining readability through the Flesch score, we also analyze the Flesch-Kincaid Grade Level (FKGL) in relation to sentiment scores for each inaugural address. The FKGL estimates the U.S. school grade level required to understand a speech, providing a direct measure of textual complexity. By comparing FKGL to sentiment scores, we explore how the sophistication of presidential language interacts with the emotional tone conveyed. Using clustering methods to group speeches based on their grade level and sentiment characteristic offering a nuanced view of how different presidents have communicated their visions to the nation.

```{r}
#| label: Plotting-Flesch-Kincaid-vs-Sentiment
#| fig-cap: "Figure [?]: This chart displays Flesch-Kincaid and sentiment scores of president speech values."

p <- ggplot(inaug_speeches_master, aes(y = Flesch.Kincaid, x = sentiment,
                                        text = paste("President:", 
                                                    president,
                                                    "<br>Year:", 
                                                    date,
                                                    "<br>Sentiment Score:", 
                                                    sentiment,
                                                    "<br>FK Score:",
                                                    Flesch.Kincaid))) +
  scale_color_manual(values = c("red", "blue", "green", "purple", "orange", "pink")) +
  geom_point(aes(color = factor(flesch_kin_sentiment_cluster))) +
  scale_x_continuous(breaks = scales:: breaks_width(25), minor_breaks = NULL) +
  labs(y = "Flesch-Kincaid Grade Level",
       x = "Sentiment Score",
        title = "Inaguaral Address Analysis",
       subtitle = "Flesch-Kincaid Grade Level vs Sentiment Score",
       color = "Cluster") +
  theme_minimal()

ggplotly(p, tooltip = "text")
```

In this example, clustering analysis using Flesch-Kincaid vs Sentiment values does not yeild any meaningful results. Clustering appears to be driven entirely by the sentiment axis and form vertical "layers" that reflect sentiment values alone. 

This suggests that the algorithm is not detecting any significant structure in the Flesch-Kincaid axis, and as a result, the clustering lacks any interpretive value.

### FKGL vs. Time

We also examine the relationship between the Flesch-Kincaid Grade Level (FKGL) and time to track how the complexity of presidential inaugural speeches has changed. Overall, we find a clear trend that speech complexity has decreased over time, with more recent presidents using simpler, more accessible language compared to earlier eras.

```{r}
#| label: Flesh-Kincaid-over-time
#| fig-cap: "Figure [?]: This chart displays Flesch-Kincaid over time illustrating the complexity of inaugural addresses."

p <- ggplot(inaug_speeches_master, aes(x = date, y = Flesch.Kincaid,
                                       text = paste("President:", 
                                                    president,
                                                    "<br>Year:", 
                                                    date,
                                                    "<br>FK Score:",
                                                    Flesch.Kincaid))
            ) +
  geom_point() +
  geom_smooth( method = "loess", se = TRUE)+
  scale_x_continuous(breaks = scales:: breaks_width(25), minor_breaks = NULL) +
  labs(title = "Inagural Speech Complexity Over Time",
    y = "Flesch-Kincaid Grade Level",
       x = "Year") +
  theme_minimal()

ggplotly(p, tooltip = "text")

```

Possible explanations include:

-   Initially, the audience of the inaugural speech was composed of congressional leaders.
-   The first inaugural speech to a public crowd was by James Monroe in 1817.
-   The growing emphasis on mass communication, as presidents began speaking to a broader and more diverse audience.
-   The influence of modern media, which favors shorter, clearer messaging that resonates quickly with the public.

## Understanding the relationship between: Flesch Reading Ease (FRE) & Flesch-Kincaid Grading Level (FKGL)

To get a better understanding of the relationship between the Flesch Reading Ease (FRE) and the Flesch-Kincaid Grade Level (FKGL), it is helpful to plot them together on a scatterplot. 

```{r}
#| label: FKGL vs FRE

ggplot(inaug_speeches_master, aes(x = Flesch, y = Flesch.Kincaid)) +
  geom_point() +
  geom_smooth( method = "loess", se = TRUE)+
  scale_x_continuous(breaks = scales:: breaks_width(10), minor_breaks = NULL) +
  labs(title = "FRE vs FKGL",
    y = "Flesch-Kincaid Grade Level",
       x = "Flesch Reading Ease") +
  theme_minimal()

```


The Flesch Reading Ease (FRE) and the Flesch-Kincaid Grade Level (FKGL) have an inverse relationship because they are based on similar textual features—sentence length and word complexity—but use opposing scales to express readability.


$$
\text{FRE} = 206.835 - 1.015 \times \left( \frac{\text{Total Words}}{\text{Total Sentences}} \right) - 84.6 \times \left( \frac{\text{Total Syllables}}{\text{Total Words}} \right)
$$
$$
\text{FKGL} = 0.39 \times \left( \frac{\text{Total Words}}{\text{Total Sentences}} \right) + 11.8 \times \left( \frac{\text{Total Syllables}}{\text{Total Words}} \right) - 15.59
$$
As sentence length and syllables per word increase, FRE scores decrease (the text becomes harder to read), while FKGL scores increase (the grade level rises). This opposing behavior is why they have an inverse relationship: when one goes up, the other tends to go down.
