---
title: "About"
author: "Rachel Hubley"
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false

# Improve digit display and NA display for kable tables
options(scipen = 1, knitr.kable.NA = "")

# Load necessary packages
library(tidyverse)
library(ggplot2)
library(plotly)

```

# About

This blog analyzes U.S. presidential inaugural speeches to track changes in language, sentiment, and readability over time. Using the AFINN sentiment lexicon, Flesch Reading Ease formula, and Flesch-Kincaid formula, we examine how emotional tone and speech complexity have shifted from 1789 to 2009.

## Methodology

## Data Collection & Wrangling Description
The president speeches were scraped from [Avalon Project at Yale Law School](https://avalon.law.yale.edu/subject_menus/inaug.asp). 

After scraping, we removed stop words, as well as "fellow", "citizens", and "people" due to repetition.

Addition, President Van Buren, President Buchanan, President Garfield, and President Coolidge are missing from the dataset due to unavailable transcripts.

For the *Sentiment Analysis*, we calculated the average sentiment of each speech to account for length of speech and to allow for fair comparisons across speeches/


The AFINN Lexicon: This Lexicon assigns identified words with a value from -5 to 5, with negative scores representing negative sentiment and positive scores representing positive sentiment.

Average Sentiment: To account for length of speech, presidents are compared by their average sentiment score (total sentiment / number of words recognized by AFINN).


## Limitations

## Future Work

## Appendix



```{r}
#| label: fig-sentiment
#| fig-cap: "AFINN Sentiment Scores of Inaugural Speeches"

```


